### 2020-04-22 serial_mli.tex
Con't on 2.3

\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}

@@

Generation and serial intervals are under-appreciated --- often treated as auxiliary variables for estimating the reproduction number $\mathcal R$ --- yet important quantities in understanding the outset of an outbreak.
Due to their similar definitions, generation and serial intervals have been often used interchangeably even though two distributions are different.
We argue that a part of the confusion can be attributed to an apparent paradox.
As the generation-interval distribution describes the renewal process of infection, it provides a link between the epidemic growth rate $r$ and the reproduction number $\mathcal R$;
likewise, the serial-interval distribution should describe the renewal process of symptomatic infection and therefore, should provide the same link between the epidemic growth rate $r$ and the reproduction number $\mathcal R$.
However, the current theory does not explain how two different distributions can give the same link between $r$ and $\mathcal R$ link.
We provide an answer to the paradox.
We show that serial intervals have to be defined with a reference perspective and a reference time;
the forward serial interval distribution during the exponential provides the correct link between $r$ and $\mathcal R$.
We also provide a heuristic way of addressing potential biases that can arise from not accounting for changes in these references.
Our study reveals gaps in the current understanding of serial intervals and provides rationale for reassessing serial intervals of COVID-19.

@@

Since the emergence of the novel coronavirus disease (COVID-19), a significant amount of research has focused on estimating its reproduction number $\mathcal R$ \citep{majumder2020early}.
The reproduction number is defined as the average number of secondary cases caused by a primary case;
the corresponding reproduction number in a fully susceptible population --- referred to as the basic reproduction number $\mathcal R_0$ --- allows us to predict the extent to which a disease will spread in the population and the amount of intervention to prevent an outbreak \citep{anderson1991infectious}.
Since reproduction number cannot be measured directly, particularly during the outset of an outbreak, it is often estimated from the observed exponential growth rate using generation- and serial-interval distributions (e.g., \cite{du2020serial, jung2020real, li2020early, zhao2020preliminary}).

The generation interval is defined as the time between when an individual (infector) is infected and when an individual infects another person (infectee);
the generation-interval distributions plays a key role in shaping the relationship between the exponential growth rate $r$ and the reproduction number $\mathcal R$ \citep{wallinga2007generation}.
Similarly, the serial interval is defined as the time between when an infector and an infectee become \emph{symptomatic} \citep{svensson2007note}.
While serial intervals are similar to generation intervals, previous studies have noted that, in many contexts, serial intervals are expected to have larger variances than generation intervals but have the same mean \citep{svensson2007note,klinkenberg2011correlation,te2013estimating,champredon2018equivalence};
some studies further suggested that using serial intervals can give different estimates of $\mathcal R$ \citep{britton2019estimation}.
Even these distributions were clearly distinguished over a decade ago \citep{svensson2007note}, 
the need for a better conceptual and theoretical framework for understanding their differences is becoming clearer as the COVID-19 pandemic unfolds:
Researchers continue to rely on both generation and serial intervals to estimate $\mathcal R$ for COVID-19 without making a clear distinction.

@@

Here, we provide an answer to this paradox, by showing that the relevant interval for the renewal framework is what is called the ``forward'' interval, and that the forward serial (but not generation) interval is affected by the rate of growth $r$ during the exponential-growth phase.
We develop a new framework for understanding serial intervals and show that the initial forward serial-interval distribution gives the correct value of $\mathcal R$.
%\jd{Come back to this. Not just over time, right?} \swp{What do you mean? Everything is about time here... you haven't read the results section so maybe that's why?}
However, using inaccurately defined serial intervals or failing to account for changes in the observed serial-interval distributions over the course of an epidemic can bias the estimate of $\mathcal R$.
We apply our framework to serial intervals of COVID-19 and lay out several principles to consider in using information about serial intervals and other epidemiological time delays in the analysis of the ongoing pandemic.

@@

We first begin by describing a general framework for characterizing a distribution of time delays between two epidemiological events;
these events can be defined either within an infected individual (e.g., infection and symptom onset of an individual, the incubation period) or between infected individuals (e.g., symptom onsets of an infector and an infectee, the serial interval).
We can further divide these events into \emph{primary} and \emph{secondary} events.
When we measure an epidemiological time delay within an infected individual (e.g., the incubation period), the primary event is the event that always or usually occurs before the secondary event ---
most epidemiological events that can be observed within an individual have clear direction (e.g., infection and onset of symptoms) but some may not (e.g., onset of infectiousness and onset of symptoms).
% \jd{Rethink, or else rethink ``any'' above: the ``hidden'' time (my term) is the time between onset of infectiousness and onset of symptoms; it is within, but can be positive or negative. I'm not saying we should use it here\ldots}
When we measure an epidemiological time delay between infected individuals (e.g., the serial interval), 
the primary and secondary events are defined in terms of the direction of transmission:
The primary event refers to the event that occurs within an infector and does not necessarily occur before the secondary event.

We model time delays between a primary and a secondary event from a cohort perspective.
A primary cohort consists of \emph{all} individuals whose primary event occurred at a given time; 
a secondary cohort is defined similarly based on the secondary events.
For example, when we are measuring serial intervals, a primary cohort $\pi$ consists of all infectors who became symptomatic at time $\pi$.
Then, for each primary cohort $\pi$, we can define the expected time distribution between primary and secondary events.
We refer to this distribution as the forward delay distribution and denote it as $f_\pi(\tau)$.

@@

The serial interval is defined as the time between when an infector becomes symptomatic and when and infectee becomes symptomatic.
Serial intervals $\tau$ have been typically written in the form:
\begin{equation}
\tau = - x_0 + \sigma + x_1
\end{equation}
where $x_0$ and $x_1$ represent the realized time from infection to symptom onset of an infector and an infectee, respectively, and $\sigma$ represents the realized generation interval.
Previous studies have often assumed that $x_0$ and $x_1$ follow the same distributions and concluded that the serial and generation intervals have the same mean \citep{svensson2007note,klinkenberg2011correlation,champredon2018equivalence, britton2019estimation}.
% \jd{I would drop the last clause. I still argue that this is true for intrinsic SIs, and we don't want to get into that here. You can just keep rolling forward at this point.}

@@

The cohort-based framework allows us to understand the distribution of serial intervals we expect to observe when incidence is changing.
Given that an infector became symptomatic at time $\pi$, we first go backward in time by asking when the infector was infected, and then go forward in time by asking first when the infectee was infected and then when the infectee became symptomatic;
this defines the forward serial interval.
In \fref{diagram}A, we see that $x_0$ represents the backward incubation period of the infector who became symptomatic at time $\pi$;
$\sigma$ represents the forward generation interval of the infector who came infected at time $\pi - x_0$;
and $x_1$ represents the forward incubation period for the infectees who became infected at time $\pi - x_0 + \sigma$.

>>>>

\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\mli}[1]{\comment{blue}{}{#1}}

@@

Generation and serial intervals are under-appreciated --- often treated as auxiliary variables for estimating the reproduction number $\mathcal R$ --- yet important quantities in understanding the outset of an outbreak.
\mli{Feels really harsh and unfair. People do care about ``time intervals''. In fact, from a selfish point of view, people care more how long they are going to be ill than how many people infect.}
Due to their similar definitions, generation and serial intervals have been often used interchangeably even though two distributions are different.
We argue that a part of the confusion can be attributed to an apparent paradox.
As the generation-interval distribution describes the renewal process of infection, it provides a link between the epidemic growth rate $r$ and the reproduction number $\mathcal R$;
likewise, the serial-interval distribution should describe the renewal process of symptomatic infection and therefore, should provide the same link between the epidemic growth rate $r$ and the reproduction number $\mathcal R$.
However, the current theory does not explain how two different distributions can give the same link between $r$ and $\mathcal R$ link.
We provide an answer to the paradox.
We show that serial intervals have to be defined with a reference perspective and a reference time;
the forward serial interval distribution during the exponential provides the correct link between $r$ and $\mathcal R$.
We also provide a heuristic way of addressing potential biases that can arise from not accounting for changes in these references.
Our study reveals gaps in the current understanding of serial intervals and provides rationale for reassessing serial intervals of COVID-19.

@@

Since the emergence of the novel coronavirus disease (COVID-19), a significant amount of research has focused on estimating its reproduction number $\mathcal R$ \citep{majumder2020early}.
\mli{Isn't this true for every disease and every outbreak? Maybe put covid19 at the end of the sentence?}
The reproduction number is defined as the average number of secondary cases caused by a primary case;
the corresponding reproduction number in a fully susceptible population --- referred to as the basic reproduction number $\mathcal R_0$ --- allows us to predict the extent to which a disease will spread in the population and the amount of intervention to prevent an outbreak \citep{anderson1991infectious}.
Since reproduction number cannot be measured directly, particularly during the outset of an outbreak, it is often estimated from the observed exponential growth rate using generation- and serial-interval distributions (e.g., \cite{du2020serial, jung2020real, li2020early, zhao2020preliminary}).
\mli{The first part of the last sentence is false, if you have contact tracing, you know the reproduction number directly. If I know exactly who and how many people I infect, that is the reproduction number by definition. Gr-link is just a way to estimate R. Probably good to explain why we like it. I.e. based on reporting data, we can estimate little r, if we don't know much about G or S, we can use some prior or similar known thing as a first approximation proxy.}

The generation interval is defined as the time between when an individual (infector) is infected and when an individual infects another person (infectee);
the generation-interval distributions plays a key role in shaping the relationship between the exponential growth rate $r$ and the reproduction number $\mathcal R$ \citep{wallinga2007generation}. \mli{also cite park et al? generation paper?}
Similarly, the serial interval is defined as the time between when an infector and an infectee become \emph{symptomatic} \citep{svensson2007note}.
While serial intervals are similar to generation intervals, previous studies have noted that, in many contexts, serial intervals are expected to have larger variances than generation intervals but have the same mean \citep{svensson2007note,klinkenberg2011correlation,te2013estimating,champredon2018equivalence};
some studies further suggested that using serial intervals can give different estimates of $\mathcal R$ \citep{britton2019estimation}.
Even these distributions were clearly distinguished over a decade ago \citep{svensson2007note}, 
the need for a better conceptual and theoretical framework for understanding their differences is becoming clearer as the COVID-19 pandemic unfolds:
Researchers continue to rely on both generation and serial intervals to estimate $\mathcal R$ for COVID-19 without making a clear distinction.

@@

Here, we provide an answer to this paradox, by showing that the relevant interval for the renewal framework is what is called the ``forward'' interval, and that the forward serial (but not generation) interval is affected by the rate of growth $r$ during the exponential-growth phase.
We develop a new framework for understanding serial intervals and show that the initial forward serial-interval distribution gives the correct value of $\mathcal R$.
%\jd{Come back to this. Not just over time, right?} \swp{What do you mean? Everything is about time here... you haven't read the results section so maybe that's why?}
However, using inaccurately defined serial intervals or failing to account for changes in the observed serial-interval distributions over the course of an epidemic can bias the estimate of $\mathcal R$.
\mli{How bad can it be? Compared to all the other problems people are sweeping under the rug (I am defintely not saying it is okay to misuse it, but it is going to change by 1\%, 10\% ...)}
We apply our framework to serial intervals of COVID-19 and lay out several principles to consider in using information about serial intervals and other epidemiological time delays in the analysis of the ongoing pandemic.

@@

We first begin by describing a general framework for characterizing a distribution of time delays between two epidemiological events;
these events can be defined either within an infected individual (e.g., infection and symptom onset of an individual, the incubation period) or between infected individuals (e.g., symptom onsets of an infector and an infectee, the serial interval).
We can further divide these events into \emph{primary} and \emph{secondary} events.
When we measure an epidemiological time delay within an infected individual (e.g., the incubation period), the primary event is the event that always or usually occurs before the secondary event ---
most epidemiological events that can be observed within an individual have clear direction (e.g., infection and onset of symptoms) but some may not (e.g., onset of infectiousness and onset of symptoms).
\mli{It sounds really confusing. Onset of symptoms appears in both clear direction and the latter (not clear direction?).}
% \jd{Rethink, or else rethink ``any'' above: the ``hidden'' time (my term) is the time between onset of infectiousness and onset of symptoms; it is within, but can be positive or negative. I'm not saying we should use it here\ldots}
When we measure an epidemiological time delay between infected individuals (e.g., the serial interval), 
the primary and secondary events are defined in terms of the direction of transmission:
The primary event refers to the event that occurs within an infector and does not necessarily occur before the secondary event.
\mli{Long text is hard to read.. I had to read it multiple times to see the difference. Maybe construct a within and between infected individuals table and have a side by side comparison.}

We model time delays between a primary and a secondary event from a cohort perspective.
A primary cohort consists of \emph{all} individuals whose primary event occurred at a given time; 
a secondary cohort is defined similarly based on the secondary events.
For example, when we are measuring serial intervals, a primary cohort $\pi$ consists of all infectors who became symptomatic at time $\pi$.
\mli{$\pi$ has different units in the previous S. (1) a primary cohort of ``individuals'' and ... time ``$\pi$''}

Then, for each primary cohort $\pi$, we can define the expected time distribution between primary and secondary events.
We refer to this distribution as the forward delay distribution and denote it as $f_\pi(\tau)$.

@@

The serial interval is defined as the time between when an infector becomes symptomatic and when and infectee becomes symptomatic.
Serial intervals $\tau$ have been typically written in the form:
\begin{equation}
\tau = - x_0 + \sigma + x_1
\end{equation}
where $x_0$ and $x_1$ represent the realized time from infection to symptom onset of an infector and an infectee, respectively, and $\sigma$ represents the realized generation interval.
\mli{This equation is too compacted. $x_0$ is the incubation period of P; $x_1$ is the incubation period of S; $GI = \sigma = x_0 + waiting_time_P$. The $x_0$ cancels out and you are left with $waiting_time_P and x_1$. Maybe use the picture from my thesis, and start from the simple construction and show how to get it to equation 3. It is not at all an obvious until I took apart the pieces to validate this is true.}
Previous studies have often assumed that $x_0$ and $x_1$ follow the same distributions and concluded that the serial and generation intervals have the same mean \citep{svensson2007note,klinkenberg2011correlation,champredon2018equivalence, britton2019estimation}.
% \jd{I would drop the last clause. I still argue that this is true for intrinsic SIs, and we don't want to get into that here. You can just keep rolling forward at this point.}

@@

The cohort-based framework allows us to understand the distribution of serial intervals we expect to observe when incidence is changing.
Given that an infector became symptomatic at time $\pi$, we first go backward in time by asking when the infector was infected, and then go forward in time by asking first when the infectee was infected and then when the infectee became symptomatic;
this defines the forward serial interval.
In \fref{diagram}A, we see that $x_0$ represents the backward incubation period of the infector who became symptomatic at time $\pi$;
$\sigma$ represents the forward generation interval of the infector who came infected at time $\pi - x_0$;
and $x_1$ represents the forward incubation period for the infectees who became infected at time $\pi - x_0 + \sigma$.
\mli{Okay, I suggest switching places... explain this first before the first paragraph. All this forward and backwards thing is making your write backwards as well. In figure 1A, calling that a forward generation interval feels weird. The focal point is in the middle of the generation interval, it is only seeing the waiting time.}
